\bf The Latent Structure Block Model

### 2024-02-22 draft/draft.Rmd
minor changes

---
title: Manifold Clustering in the Generalized Random Dot Product Graph

@@

bibliography: bibliography.bib
output: rticles::asa_article
urlcolor: blue
header-includes:
- \usepackage{float}
- \usepackage{mathtools}
- \usepackage{natbib}
- \usepackage[linesnumbered,ruled,vlined]{algorithm2e} 
- \usepackage{verbatim}
- \usepackage{amsthm}
- \usepackage{comment}
- \usepackage{amsfonts}
---

@@

One common analysis task for graph and network data is community detection, which assumes that each vertex of a graph has a hidden community label. The goal of the analysis is then to estimate these labels upon observing a graph. 
In order to perform this analysis as a statistical inference task is to define a probability model with inherent community structure. We call such models *block models*: 
First, each vertex is assigned a label $z_1, ..., z_n \in \{1, 2, ..., K\}$ where $K \ll n$. 
Then each edge probability $P_{ij}$ is said to depend on the labels $z_i$ and $z_j$, possibly along with some other parameters. 
For example, the stochastic block model (SBM) sets a fixed edge probability for each pair of communities, i.e., $P_{ij} = \omega_{z_i, z_j}$. 
The degree-corrected block model (DCBM) assigns an additional parameter $\theta_i$ to each vertex by which edge probabilities are scaled, i.e., $P_{ij} = \theta_i \theta_j \omega_{z_i, z_j}$. 
The popularity adjusted block model (PABM) assigns $K$ parameters to each vertex $\lambda_{i1}, \lambda_{i2}, ..., \lambda_{iK}$ that describe that vertex's affinity toward each community; 
the edge probability between vertices $i$ and $j$ is then defined as the product of vertex $i$'s affinity toward vertex $j$'s community and vertex $j$'s affinity toward vertex $i$'s community, i.e., $P_{ij} = \lambda_{i z_j} \lambda_{j z_i}$.

The three block model types, as well as the $\ER$ model, impose structure on $P$, including on the rank of $P$. 
$P$ has rank 1 for the $\ER$ model, rank $K$ for the SBM and DCBM, and rank $K^2$ for the PABM. 
This provides the intuition behind another family of Bernoulli graphs called the *random dot product graph* (RDPG) and *generalized random dot product graph* (GRDPG). 
In the RDPG, each vertex has a corresponding latent vector in $d$-dimensional Euclidean space, where $d$ is the rank of $P$ and $P$ is positive semidefinite. Then the edge probability between each pair of vertices is defined as the inner product between the corresponding latent vectors, i.e., $P_{ij} = x_i^\top x_j$. If the latent vectors are collected in a data matrix $X = \bigl[ x_1 \mid \cdots \mid x_n \bigr]^\top$, then the edge probability matrix for the RDPG is $P = X X^\top$. 
Similarly, the edge probability between each pair of vertices for the GRDPG is defined as the indefinite inner product between the corresponding latent vectors, i.e., $P_{ij} = x_i^\top I_{p,q} x_j$, where $I_{p,q} = \blockdiag(I_p, -I_q)$ and $p + q = d$. Then the edge probability matrix for the GRDPG is $P = X I_{p,q} X^\top$. This allows for a model similar to the RDPG for non-positive semidefinite $P$. 
While the RDPG and GRDPG do not necessarily have community structure, it has been shown that block models are specific cases of the RDPG or GRDPG in which latent vectors are organized by community. This includes the SBM, in which communities correspond to point masses, DCBM, in which communities correspond to line segments, and PABM, in which communities correspond to orthogonal subspaces. In this work, we extend this idea to communities organized into more general latent structures. In particular, we assume that each community corresponds to a manifold in the latent space.

@@

All Bernoulli graphs are generalized random dot product graphs. 
Whether this is useful for inference depends on the structure of the latent space. 
In the case of the $\ER$ model, SBM, DCBM, and PABM, the latent structure is linear, and the linearity can be exploited for community detection and parameter estimation. 
In this section, we discuss general, often nonlinear latent structure models, focusing on those with community structure.

@@

\begin{algorithm}[h]
\DontPrintSemicolon
\SetAlgoLined
\KwData{Adjacency matrix $A$, number of communities $K$, embedding dimensions $p$, $q$, stopping criterion $\epsilon$, $m_k \leq n_k$ known community assignments for each community}
\KwResult{Community assignments $1, ..., K$, curves $g_1, ..., g_K$}
Compute $X$, the ASE of $A$ using the $p$ most positive and $q$ most negative eigenvalues and their corresponding eigenvectors.\;
Fit curves $g_1, ..., g_K$ using each of the $m_1, ..., m_K$ points with known community labels by minimizing $\sum_{j=1}^{m_i} \|x_j - g_k(t_j)\|^2$.\;
Assign labels $z_1, ..., z_n$ to each $x_1, ..., x_n$ by minimizing $\|x_i - g_k(t_i)\|^2$ for $k$, holding the initial known labels constant.\; 
\Repeat {the change in $\sum_k \sum_{i \in C_k} \|x_i - g_k(t_i)\|^2$ is less than $\epsilon$} {
\For {$k = 1, ..., K$} {
Define $X_k$ as the rows of $X$ for which $z_i = k$.\;
Fit curve $g_k$ and positions $t_{k_i}$ to $X_k$ by minimizing $\sum_{k_i} \|x_{k_i} - g_k(t_{k_i})\|^2$.\;
}
\For {$k = 1, ..., K$} {
Assign $z_i \leftarrow \arg\min_\ell \|x_i - g_\ell(t_i)\|^2$, holding the known initial labels constant.\
}
}
\caption{Semi-supervised $K$-curves clustering.}
\end{algorithm}

@@

```{r}
clustering.df <- readr::read_csv('../simulations/balanced.csv')
```

@@

>>>>

---
title: The Latent Structure Block Model

@@

bibliography: bibliography.bib
output: rticles::asa_article
# urlcolor: blue
header-includes:
- \usepackage{float}
- \usepackage{mathtools}
- \usepackage{natbib}
- \usepackage[linesnumbered,ruled,vlined]{algorithm2e} 
- \usepackage{verbatim}
- \usepackage{amsthm}
- \usepackage{comment}
- \usepackage{amsfonts}
- \usepackage{pdfpages}
- \usepackage{xcolor}
---

@@

One common analysis task for graph and network data is community detection, which assumes that each vertex of a graph has a hidden community label. 
The goal of the analysis is then to estimate these labels upon observing a graph. 
One approach is to define a random graph model that includes community labels. 
First, each vertex is assigned a label $z_1, ..., z_n \in \{1, 2, ..., K\}$ where $K \ll n$. 
Then the edge probabilities are restricted such that they depend on the labels of the vertices, and possibly some other parameters specific to each vertex, i.e., $P_{ij} = g(z_i, z_j, \phi_i, \phi_j)$, where $\phi_i$ is a set of parameters corresponding to vertex $i$. 
We call such models *block models*. 
The simplest and best known block model is the stochastic block model (SBM), which sets a fixed edge probability for each pair of communities, i.e., $P_{ij} = \theta_{z_i, z_j}$. 
The degree-corrected block model (DCBM) assigns an additional parameter $\omega_i$ to each vertex by which edge probabilities are scaled, i.e., $P_{ij} = \omega_i \omega_j \theta_{z_i, z_j}$. 
The popularity adjusted block model (PABM) assigns $K$ parameters to each vertex $\lambda_{i1}, \lambda_{i2}, ..., \lambda_{iK}$ that describe that vertex's affinity toward each community; 
the edge probability between vertices $i$ and $j$ is then defined as the product of vertex $i$'s affinity toward vertex $j$'s community and vertex $j$'s affinity toward vertex $i$'s community, i.e., $P_{ij} = \lambda_{i z_j} \lambda_{j z_i}$. 
The three block model types, as well as the $\ER$ model, impose structure on $P$, including on the rank of $P$. 
$P$ has rank 1 for the $\ER$ model, rank $K$ for the SBM and DCBM, and rank $K^2$ for the PABM. 
This provides the intuition behind another family of Bernoulli graphs called the *random dot product graph* (RDPG) and *generalized random dot product graph* (GRDPG). 
In the RDPG, each vertex has a corresponding latent vector in $d$-dimensional Euclidean space, where $d$ is the rank of $P$ and $P$ is positive semidefinite. 
Then the edge probability between each pair of vertices is defined as the inner product between the corresponding latent vectors, i.e., $P_{ij} = x_i^\top x_j$. If the latent vectors are collected in a data matrix $X = \bigl[ x_1 \mid \cdots \mid x_n \bigr]^\top$, then the edge probability matrix for the RDPG is $P = X X^\top$. 
Similarly, the edge probability between each pair of vertices for the GRDPG is defined as the indefinite inner product between the corresponding latent vectors, i.e., $P_{ij} = x_i^\top I_{p,q} x_j$, where $I_{p,q} = \blockdiag(I_p, -I_q)$ and $p + q = d$. Then the edge probability matrix for the GRDPG is $P = X I_{p,q} X^\top$. 
This allows for a model similar to the RDPG for non-positive semidefinite $P$. 
While the RDPG and GRDPG do not necessarily have community structure, it has been shown that block models are specific cases of the RDPG or GRDPG in which latent vectors are organized by community. 
This includes the SBM, in which communities correspond to point masses, DCBM, in which communities correspond to line segments, and PABM, in which communities correspond to orthogonal subspaces. 
In this work, we extend this idea to communities organized into more general latent structures. In particular, we assume that each community corresponds to a manifold in the latent space.

@@

All Bernoulli graphs are generalized random dot product graphs. 
For a specific example, consider the two-community DCBM and its GRDPG latent space representation. 
In the DCBM, the edge probability between each pair of vertices is the product of the baseline community-wise edge probability and the degree correction factors of the vertices, or $P_{ij} = \theta_{z_i, z_j} \omega_i \omega_j$.

@@

\begin{algorithm}[H]
\DontPrintSemicolon
\SetAlgoLined
\KwData{Adjacency matrix $A$, number of communities $K$, embedding dimensions $p$, $q$, stopping criterion $\epsilon$, $m_k \leq n_k$ known community assignments for each community}
\KwResult{Community assignments $1, ..., K$, curves $g_1, ..., g_K$}
Compute $X$, the ASE of $A$ using the $p$ most positive and $q$ most negative eigenvalues and their corresponding eigenvectors.\;
Fit curves $g_1, ..., g_K$ using each of the $m_1, ..., m_K$ points with known community labels by minimizing $\sum_{j=1}^{m_i} \|x_j - g_k(t_j)\|^2$.\;
Assign labels $z_1, ..., z_n$ to each $x_1, ..., x_n$ by minimizing $\|x_i - g_k(t_i)\|^2$ for $k$, holding the initial known labels constant.\; 
\Repeat {the change in $\sum_k \sum_{i \in C_k} \|x_i - g_k(t_i)\|^2$ is less than $\epsilon$} {
\For {$k = 1, ..., K$} {
Define $X_k$ as the rows of $X$ for which $z_i = k$.\;
Fit curve $g_k$ and positions $t_{k_i}$ to $X_k$ by minimizing $\sum_{k_i} \|x_{k_i} - g_k(t_{k_i})\|^2$.\;
}
\For {$k = 1, ..., K$} {
Assign $z_i \leftarrow \arg\min_\ell \|x_i - g_\ell(t_i)\|^2$, holding the known initial labels constant.\
}
}
\caption{Semi-supervised $K$-curves clustering.}
\end{algorithm}

@@

```{r}
clustering.df <- readr::read_csv('../simulations/balanced-2/balanced.csv')
```

@@

# References

### 2024-02-22 draft/draft.tex
minor changes

% !TeX program = pdfLaTeX
\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{textcomp}
\usepackage[hyphens]{url} % not crucial - just used below for the URL
\usepackage{hyperref}

@@

% DON'T change margins - should be 1 inch all around.
\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-.5in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.3in}%
\addtolength{\topmargin}{-.8in}%

@@

\usepackage{float}
\usepackage{mathtools}
\usepackage{natbib}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage{verbatim}
\usepackage{amsthm}
\usepackage{comment}
\usepackage{amsfonts}

@@

\if0\blind
{
  \title{\bf Manifold Clustering in the Generalized Random Dot Product
Graph}

@@

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Manifold Clustering in the Generalized Random Dot Product
Graph}
  \end{center}
  \medskip
} \fi

@@

\noindent%
{\it Keywords:} block models, community detection, coordinate descent,
latent structure models, manifold clustering, random dot product graph
\vfill

\newpage
\spacingset{1.45} % DON'T change the spacing!

@@

One common analysis task for graph and network data is community
detection, which assumes that each vertex of a graph has a hidden
community label. The goal of the analysis is then to estimate these
labels upon observing a graph. In order to perform this analysis as a
statistical inference task is to define a probability model with
inherent community structure. We call such models \emph{block models}:
First, each vertex is assigned a label
\(z_1, ..., z_n \in \{1, 2, ..., K\}\) where \(K \ll n\). Then each edge
probability \(P_{ij}\) is said to depend on the labels \(z_i\) and
\(z_j\), possibly along with some other parameters. For example, the
stochastic block model (SBM) sets a fixed edge probability for each pair
of communities, i.e., \(P_{ij} = \omega_{z_i, z_j}\). The
degree-corrected block model (DCBM) assigns an additional parameter
\(\theta_i\) to each vertex by which edge probabilities are scaled,
i.e., \(P_{ij} = \theta_i \theta_j \omega_{z_i, z_j}\). The popularity
adjusted block model (PABM) assigns \(K\) parameters to each vertex
\(\lambda_{i1}, \lambda_{i2}, ..., \lambda_{iK}\) that describe that
vertex's affinity toward each community; the edge probability between
vertices \(i\) and \(j\) is then defined as the product of vertex
\(i\)'s affinity toward vertex \(j\)'s community and vertex \(j\)'s
affinity toward vertex \(i\)'s community, i.e.,
\(P_{ij} = \lambda_{i z_j} \lambda_{j z_i}\).

The three block model types, as well as the
\(\text{Erd\"{o}s-R\'{e}nyi}\) model, impose structure on \(P\),
including on the rank of \(P\). \(P\) has rank 1 for the
\(\text{Erd\"{o}s-R\'{e}nyi}\) model, rank \(K\) for the SBM and DCBM,
and rank \(K^2\) for the PABM. This provides the intuition behind
another family of Bernoulli graphs called the \emph{random dot product
graph} (RDPG) and \emph{generalized random dot product graph} (GRDPG).
In the RDPG, each vertex has a corresponding latent vector in
\(d\)-dimensional Euclidean space, where \(d\) is the rank of \(P\) and
\(P\) is positive semidefinite. Then the edge probability between each
pair of vertices is defined as the inner product between the
corresponding latent vectors, i.e., \(P_{ij} = x_i^\top x_j\). If the
latent vectors are collected in a data matrix
\(X = \bigl[ x_1 \mid \cdots \mid x_n \bigr]^\top\), then the edge
probability matrix for the RDPG is \(P = X X^\top\). Similarly, the edge
probability between each pair of vertices for the GRDPG is defined as
the indefinite inner product between the corresponding latent vectors,
i.e., \(P_{ij} = x_i^\top I_{p,q} x_j\), where
\(I_{p,q} = \mathrm{blockdiag}(I_p, -I_q)\) and \(p + q = d\). Then the
edge probability matrix for the GRDPG is \(P = X I_{p,q} X^\top\). This
allows for a model similar to the RDPG for non-positive semidefinite
\(P\). While the RDPG and GRDPG do not necessarily have community
structure, it has been shown that block models are specific cases of the
RDPG or GRDPG in which latent vectors are organized by community. This
includes the SBM, in which communities correspond to point masses, DCBM,
in which communities correspond to line segments, and PABM, in which
communities correspond to orthogonal subspaces. In this work, we extend
this idea to communities organized into more general latent structures.
In particular, we assume that each community corresponds to a manifold
in the latent space.

@@

All Bernoulli graphs are generalized random dot product graphs. Whether
this is useful for inference depends on the structure of the latent
space. In the case of the \(\text{Erd\"{o}s-R\'{e}nyi}\) model, SBM,
DCBM, and PABM, the latent structure is linear, and the linearity can be
exploited for community detection and parameter estimation. In this
section, we discuss general, often nonlinear latent structure models,
focusing on those with community structure.

@@

\bibliographystyle{agsm}
\bibliography{bibliography.bib}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
% !TeX program = pdfLaTeX
\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
\usepackage[]{natbib}
\usepackage{textcomp}

@@

% DON'T change margins - should be 1 inch all around.
\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-1in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.7in}%
\addtolength{\topmargin}{-1in}%

@@

\usepackage{float}
\usepackage{mathtools}
\usepackage{natbib}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage{verbatim}
\usepackage{amsthm}
\usepackage{comment}
\usepackage{amsfonts}
\usepackage{pdfpages}
\usepackage{xcolor}

\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\hypersetup{
  pdftitle={The Latent Structure Block Model},
  pdfkeywords={block models, community detection, coordinate descent,
latent structure models, manifold clustering, random dot product graph},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

\if0\blind
{
  \title{\bf The Latent Structure Block Model}

@@

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf The Latent Structure Block Model}
  \end{center}
  \medskip
} \fi

@@

\noindent%
{\it Keywords:} block models, community detection, coordinate descent,
latent structure models, manifold clustering, random dot product graph

\vfill

\newpage
\spacingset{1.9} % DON'T change the spacing!

@@

One common analysis task for graph and network data is community
detection, which assumes that each vertex of a graph has a hidden
community label. The goal of the analysis is then to estimate these
labels upon observing a graph. One approach is to define a random graph
model that includes community labels. First, each vertex is assigned a
label \(z_1, ..., z_n \in \{1, 2, ..., K\}\) where \(K \ll n\). Then the
edge probabilities are restricted such that they depend on the labels of
the vertices, and possibly some other parameters specific to each
vertex, i.e., \(P_{ij} = g(z_i, z_j, \phi_i, \phi_j)\), where \(\phi_i\)
is a set of parameters corresponding to vertex \(i\). We call such
models \emph{block models}. The simplest and best known block model is
the stochastic block model (SBM), which sets a fixed edge probability
for each pair of communities, i.e., \(P_{ij} = \theta_{z_i, z_j}\). The
degree-corrected block model (DCBM) assigns an additional parameter
\(\omega_i\) to each vertex by which edge probabilities are scaled,
i.e., \(P_{ij} = \omega_i \omega_j \theta_{z_i, z_j}\). The popularity
adjusted block model (PABM) assigns \(K\) parameters to each vertex
\(\lambda_{i1}, \lambda_{i2}, ..., \lambda_{iK}\) that describe that
vertex's affinity toward each community; the edge probability between
vertices \(i\) and \(j\) is then defined as the product of vertex
\(i\)'s affinity toward vertex \(j\)'s community and vertex \(j\)'s
affinity toward vertex \(i\)'s community, i.e.,
\(P_{ij} = \lambda_{i z_j} \lambda_{j z_i}\). The three block model
types, as well as the \(\text{Erd\"{o}s-R\'{e}nyi}\) model, impose
structure on \(P\), including on the rank of \(P\). \(P\) has rank 1 for
the \(\text{Erd\"{o}s-R\'{e}nyi}\) model, rank \(K\) for the SBM and
DCBM, and rank \(K^2\) for the PABM. This provides the intuition behind
another family of Bernoulli graphs called the \emph{random dot product
graph} (RDPG) and \emph{generalized random dot product graph} (GRDPG).
In the RDPG, each vertex has a corresponding latent vector in
\(d\)-dimensional Euclidean space, where \(d\) is the rank of \(P\) and
\(P\) is positive semidefinite. Then the edge probability between each
pair of vertices is defined as the inner product between the
corresponding latent vectors, i.e., \(P_{ij} = x_i^\top x_j\). If the
latent vectors are collected in a data matrix
\(X = \bigl[ x_1 \mid \cdots \mid x_n \bigr]^\top\), then the edge
probability matrix for the RDPG is \(P = X X^\top\). Similarly, the edge
probability between each pair of vertices for the GRDPG is defined as
the indefinite inner product between the corresponding latent vectors,
i.e., \(P_{ij} = x_i^\top I_{p,q} x_j\), where
\(I_{p,q} = \mathrm{blockdiag}(I_p, -I_q)\) and \(p + q = d\). Then the
edge probability matrix for the GRDPG is \(P = X I_{p,q} X^\top\). This
allows for a model similar to the RDPG for non-positive semidefinite
\(P\). While the RDPG and GRDPG do not necessarily have community
structure, it has been shown that block models are specific cases of the
RDPG or GRDPG in which latent vectors are organized by community. This
includes the SBM, in which communities correspond to point masses, DCBM,
in which communities correspond to line segments, and PABM, in which
communities correspond to orthogonal subspaces. In this work, we extend
this idea to communities organized into more general latent structures.
In particular, we assume that each community corresponds to a manifold
in the latent space.

@@

All Bernoulli graphs are generalized random dot product graphs. For a
specific example, consider the two-community DCBM and its GRDPG latent
space representation. In the DCBM, the edge probability between each
pair of vertices is the product of the baseline community-wise edge
probability and the degree correction factors of the vertices, or
\(P_{ij} = \theta_{z_i, z_j} \omega_i \omega_j\).

@@

\bibliographystyle{plain}
\renewcommand\refname{References}
\bibliography{bibliography.bib}
